|

### 2021-04-13 paper/v2_taylor_francis/broadband.speed.covid.Rmd

---
title: |
  Working from home and digital divides: resilience during the pandemic
author:
  - name: Hannah Budnitz
    affil: a
  - name: Emmanouil Tranos
    affil: b, c
    email: e.tranos@bristol.ac.uk
affiliation:
  - num: a
    address: |
      Transport Studies Unit, University of Oxford, UK
  - num: b
    address: |
      School of Geographic Sciences, University of Bristol, UK
  - num: c
    address: |
      The Alan Turing Institute, UK
bibliography: bibliography.bib
abstract: |
  This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides. Using the UK as a case study, this paper studies how the quality and reliability of internet services, as reflected in *experienced* internet upload speeds during the spring 2020 lockdown, may reinforce or redress the spatial and social dimensions of digital divisions. Fast, reliable internet connections are necessary for the population to be able to work from home. Although not every place hosts individuals in occupations which allow for telecommuting nor with the necessary skills to effectively use the internet to telecommute, good internet connectivity is also essential to local economic resilience in a period like the current pandemic. Employing data on individual broadband speed tests and state-of-the-art time-series clustering methods, we create clusters of UK local authorities with similar temporal signatures of experienced upload speeds. We then associate these clusters of local authorities with their socioeconomic and geographic characteristics to explore how they overlap with or diverge from the existing economic and digital geography of the UK. Our analysis enables us to better understand how the spatial and social distribution of both occupations and online accessibility intersect to enable or hinder the practice of telecommuting at a time of extreme demand.
keywords: |
  covid; internet; working from home; broadband speed; time-series clusters
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{dcolumn}
  \def\tightlist{}
output: rticles::tf_article
---

@@

<!-- WFH and digital divides-->
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The capability to telecommute is related to all three levels of digital divides,
but more importantly leads to differentiated outcomes regarding the economic
resilience of people and places to overcome a systemic shock such as the current
pandemic.

Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
In some aspects, these returns may have increased during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with access to and quality of internet
connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies. Rural areas, predictably, fare worse.
This finding broadly applies in the UK, where studies that also analysed broadband 
speed checks concluded that average speeds are lower in rural areas, something that
has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
In contrast, @riddlesden2014broadband found that levels of deprivation did not 
correlate with first level digital divides.

However, as @blank2018local highlight, variation in individual internet 
availability and uptake is a product of more complex spatial and 
demographic characteristics than simple rurality or urbanisation.
Dense urban areas were shown to suffer more from slowdown during peak hours, 
although these services were more likely to be improved by increased competition 
between providers, such as between new entrants and Virgin Media cable connections
[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
available to only $45$% of premises in the UK [@ofcom2016], where the more
lucrative and competitive market originally attracted the cable TV provider. 
Whether the variation in infrastructure quality and reliability affects the 
spatial footprint of telecommuting has not previously been investigated, in part 
because telecommuting has not previously been a cause of concentrated demand and 
pressure on internet services.

@@

The extreme demand during the pandemic thus provides a new opportunity to understand
how internet infrastructure quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive 
video conferencing required in order to avoid the face-to-face contact that could 
increase the spread of infection.
We seek to answer how internet service resilience could contribute to or reduce 
economic resilience when the latter is dependent upon the capability to 
work from home.
We also aim to improve our understanding of the impact of first level digital
divisions on telecommuting, and whether this results in more fundamental third
level digital divisions than has previously been perceived.

@@

<!-- WFH broader consequences-->
Therefore, home-based digital infrastructure which considers upload speeds and 
working day reliability as well as availability are likely to be particularly 
important in a future where telecommuting might be a more common means of accessing 
work and broadband services must be fit for purpose. 
Although the long-term effects of such drastic changes in telecommuting and 
attitudes towards working from home are difficult to predict, the reliability of 
home broadband services deserve more consideration than in the past. 
This would represent a switch from previous demand-side broadband policies in the UK
which tended to be aimed at supporting small and medium enterprises [@HENDERSON2020102024].
SUch policies were based on previous research regarding the productivity effects 
of broadband infrastructure [@DESTEFANO2018110]. However, this stream of research 
tended not to consider residential locations as places that host economic activity.
Such policies also made some assumptions about the quality of 
service in residential markets that could be assured without government intervention 
-- an assumption which our analysis suggest is not entirely accurate.

>>>>

---
title: |
  Working from home and digital divides: resilience during the pandemic
author:
  - name: Hannah Budnitz
    affil: a
    email: hannah.budnitz@ouce.ox.ac.uk
  - name: Emmanouil Tranos
    affil: b, c, \dagger
    email: e.tranos@bristol.ac.uk
affiliation:
  - num: a
    address: |
      Transport Studies Unit, University of Oxford, UK
  - num: b
    address: |
      School of Geographic Sciences, University of Bristol, UK
  - num: c
    address: |
      The Alan Turing Institute, UK
  - num: \dagger
    address: |
      Corresponding author
bibliography: bibliography.bib
abstract: |
  This paper offers a new perspective on telecommuting from the viewpoint of the complex web of digital divides. Using the UK as a case study, this paper studies how the quality and reliability of internet services, as reflected in *experienced* internet upload speeds during the spring 2020 lockdown, may reinforce or redress the spatial and social dimensions of digital divisions. Fast, reliable internet connections are necessary for the population to be able to work from home. Although not every place hosts individuals in occupations which allow for telecommuting nor with the necessary skills to effectively use the internet to telecommute, good internet connectivity is also essential to local economic resilience in a period like the current pandemic. Employing data on individual broadband speed tests and state-of-the-art time-series clustering methods, we create clusters of UK local authorities with similar temporal signatures of experienced upload speeds. We then associate these clusters of local authorities with their socioeconomic and geographic characteristics to explore how they overlap with or diverge from the existing economic and digital geography of the UK. Our analysis enables us to better understand how the spatial and social distribution of both occupations and online accessibility intersect to enable or hinder the practice of telecommuting at a time of extreme demand.
keywords: |
  covid; internet; working from home; broadband speed; time-series clusters
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{dcolumn}
  \def\tightlist{}
output: rticles::tf_article
---

@@

<!-- WFH and digital divides-->
The literature describes first level digital divides in terms of the
availability and quality of internet connectivity, such as that manifest in
different geographies in the UK [@riddlesden2014broadband; @philip2017digital].
Second level digital divides consider the presence or lack of the necessary skills
to effectively utilise digital technologies and the internet
[@blank2014dimensions; @van2011internet].
The third level focuses on the heterogenous returns of internet usage among
different socioeconomic groups and, consequently, how digital technologies can
assist in bridging or further enhancing existing socioeconomic divides.
[@stern2009levels; @van2014digital; @van2015third].
The potential for these different levels of digital divide to interact is growing in 
parallel with the digitisation of various sectors of the economy and society.
For example, @li2020challenges highlights how the intersection of new demands upon
ICT infrastructure and a paucity of end users means that some areas are being left
behind in the roll-out of smart applications and services. They argue that this 
could create a geospatial 'smart divide', with areas unable to capitalise on the 
opportunities offered by such applications and services.
Likewise, the capability to telecommute is related to all three levels of 
digital divides, but more importantly leads to differentiated outcomes regarding 
the economic resilience of people and places to overcome a systemic shock such as 
the current pandemic.

Our understanding of telecommuting as a product of enabled occupations can be
described as a manifestation of the third level digital divide, as those who are able
to use digital technologies to work from home benefit from a high rate of return on
their use of the internet in terms of autonomy, flexibility, and time saved from
commuting [@peters2004employees; @siha2006telecommuting; @singh2013modeling].
In some aspects, these returns may have increased during the Covid-19 crisis,
when those with the capability to telecommute also have the ability to maintain
their employment whilst protecting their health.
However, the success of these arrangements has been dependent upon the first
level digital divide, which is associated with geospatial access to and quality 
of internet connectivity.
@SALEMINK2017360 provides a systematic review of the pre-pandemic,
first level digital divide in infrastructure quality between urban and rural areas
in various advanced economies. Rural areas, predictably, fare worse.
This finding broadly applies in the UK, where studies that also analysed broadband 
speed checks concluded that average speeds are lower in rural areas, something that
has not been improved by policy measures to increase competition [@riddlesden2014broadband; @nardotto2015unbundling].
In contrast, @riddlesden2014broadband found that levels of deprivation did not 
correlate with first level digital divides, perhaps because many deprived areas 
benefit from spillover effects of quality installed infrastructure nearby.

However, as @blank2018local highlight, variation in individual internet 
availability and uptake is a product of more complex geospatial and 
demographic characteristics than simple rurality, urbanisation or levels of deprivation.
Dense urban areas were shown to suffer more from slowdown during peak hours, 
although these services were more likely to be improved by increased competition 
between providers, such as between new entrants and Virgin Media cable connections
[@riddlesden2014broadband; @nardotto2015unbundling]. The latter were historically 
available to only $45$% of premises in the UK [@ofcom2016], where the more
lucrative and competitive market originally attracted the cable TV provider. 
Whether the geospatial variation in infrastructure quality and reliability affects 
the spatial footprint of telecommuting has not previously been investigated, 
in part because telecommuting has not previously been a cause of concentrated 
demand and pressure on internet services.

@@

The extreme demand during the pandemic thus provides a new opportunity to understand
how internet infrastructure quality, and reliability affects telecommuting, 
particularly in light of the high volumes of bandwidth-intensive 
video conferencing required in order to avoid the face-to-face contact that could 
increase the spread of infection.
We seek to answer how internet service resilience could contribute to or reduce 
economic resilience when the latter is dependent upon the capability to 
work from home.
We also aim to improve our understanding of the impact of first level digital
divisions on telecommuting, in terms of both geospatial and temporal quality of 
service, and whether this results in more fundamental third level digital divisions 
between local geographies than has previously been perceived.

@@

<!-- WFH broader consequences-->
Therefore, home-based digital infrastructure which considers upload speeds and 
working day reliability as well as availability are likely to be particularly 
important in a future where telecommuting might be a more common means of accessing 
work and broadband services must be fit for purpose. 
Although the long-term effects of such drastic changes in telecommuting and 
attitudes towards working from home are difficult to predict, the reliability of 
home broadband services deserve more consideration than in the past. 
This would represent a switch from previous demand-side broadband policies in the UK
which tended to be aimed at supporting small and medium enterprises [@HENDERSON2020102024].
Such policies were based on previous research regarding the productivity effects 
of broadband infrastructure [@DESTEFANO2018110]. However, this stream of research 
tended not to consider residential locations as places that host economic activity.
Such policies also made some assumptions about the quality of 
service in residential markets that could be assured without government intervention 
-- an assumption which our analysis suggest is not entirely accurate.
